\documentclass{article}

\usepackage{../styles/PRIMEarxiv} % imports style
\input{../shared/package} % imports packages 

% imports code formatting used across the document 
\input{../styles/rust-code-style.tex}


%Formating code listings
% https://nasa.github.io/nasa-latex-docs/html/examples/listing.html
\usepackage{listings}
\lstset
{ %Formatting for code in appendix
    basicstyle=\footnotesize,
    numbers=left,
    stepnumber=1,
    showstringspaces=false,
    tabsize=1,
    breaklines=true,
    breakatwhitespace=false,
}

%
% How to win a best paper award 
% (or, an opinionated take on how to do important research that matters)
% https://nicholas.carlini.com/writing/2026/how-to-win-a-best-paper-award.html
%

%Header
\pagestyle{fancy}
\thispagestyle{empty}
\rhead{ \textit{ }}

% Update your Headers here
\fancyhead[LO]{The Effect Propagation Process (EPP): An Axiomatic Foundation for Dynamic Causality}

%% Title
\title{The Effect Propagation Process (EPP):\newline An Axiomatic Foundation for Dynamic Causality}

%% Author
\author{
    Marvin Hansen \\
    \texttt{marvin.hansen@gmail.com} \\
    Date: \today
}


\begin{document}
    \maketitle

    \begin{abstract}
        Contemporary frameworks for computational causality provide a formal basis for inference within systems governed by static causal structures. These models, however, are predicated on assumptions of a fixed background spacetime and linear temporal progression, which become fundamental limitations when addressing complex dynamic systems where the causal laws themselves can evolve.

        This monograph introduces the Effect Propagation Process (EPP), a formal meta-calculus for dynamic causality. The EPP rests on a single axiomatic foundation: causality as spacetime-agnostic monadic functional dependency. Monadic composition supplies the directional form of a causal process, while a definitional postulate of mechanism autonomy, that the composed functions are self-contained and independently disruptable mechanisms, supplies the content that separates a causal dependency from arbitrary composition. The classical definition is recovered as a parametric specialization in which the monadic context parameters are set to the unit type and three further restrictions on the carried function are imposed; under the same restriction the intervention operator recovers Pearl's do-operator, and contrasting an intervened process against the retained factual process yields the unit-level counterfactual, the third rung of Pearl's ladder. The axiom necessitates a set of computable primitives that operationalize dynamic causality.

        The Context is an explicit and dynamic hypergraph that models the operational environment, supporting non-Euclidean and non-linear temporal structures. The Causaloid is an isomorphic recursive causal structure that enables multi-model causal reasoning over singleton, collection, or hypergraphs of causaloids.
        The CausalMonad provides linear monadic composition with carried state, configuration, error, and audit log. Both the Causaloid and the CausalMonad emit the same PropagatingEffect, and as a consequence structural causal reasoning over hypergraphs and linear functional composition compose within a single causal model. The Effect Ethos is a deontic guardrail that uses a defeasible calculus to ensure a system's actions remain verifiably aligned with its safety and mission objectives.

        The EPP is fully implemented in the open-source DeepCausality project as a layered scientific-computing stack of twenty crates that compose through higher-kinded types and share a unified metric specification, with a causal discovery language closing the loop from observational data to executable causal model. Together, the EPP and its implementation in DeepCausality provide a foundation for the design, simulation, and implementation of a new generation of robust, context-aware, and verifiably aligned dynamic causal systems.
    \end{abstract}


    \newpage

%% Table of Contents
    \tableofcontents

    \newpage

%% Introduction
    \input{introduction.tex}

%% Related Work
    \input{../epp_chapters/related_work.tex}

%% History
    \input{../epp_chapters/history.tex}

%% Effect Propagation Process
    \input{../epp_chapters/causality_as_epp.tex}

%% Metaphysics
    \input{../epp_chapters/metaphysics.tex}

%% Epistemology
    \input{../epp_chapters/epistemology.tex}

%% Teleology
    \input{../epp_chapters/teleology.tex}

%% Ontology
    \input{../epp_chapters/ontology.tex}

%% Formalization  
    \input{../epp_chapters/formalization.tex}

%% Validity 
%%\input{../epp_chapters/validity.tex}

%% Implementation 
    \input{../epp_chapters/implementation.tex}


%% Examples
    \input{../epp_chapters/examples.tex}

%% Future Work
    \input{../epp_chapters/future_work.tex}

%% Epilogue
    \input{../epp_chapters/epilogue.tex}

%% Appendix A
    \input{../epp_appendices/appendix_a.tex}

%% Glossary
    \input{../epp_chapters/glossary.tex}

%% Bibliography
    \newpage
    \bibliographystyle{unsrt}
    \addcontentsline{toc}{section}{References} % Add it to the Table of Contents
    \bibliography{../epp_bib/references}

\end{document}
